A Test of the Frequency Independence Assumption of Power System Parameters used in Geomagnetically Induced Current Estimates

### 2019-03-29 paper.tex
Updates from Overleaf

\begin{keypoints}
\item Power system parameters derived from a time series of GIC and geoelectric field measurements are frequency dependent.
\item A GIC prediction model with frequency dependence significantly outperforms a frequency independent model.
\item A model derived using GIC/geomagnetic field data provides better predictions than a GIC/geoelectric field-based model.
\end{keypoints}

\begin{abstract}
A common assumption used in estimating geomagnetically induced currents (GICs) in a power system given a nearby direct measurement or indirect estimate (via a transfer function with input of the geomagnetic field) of the horizontal geoelectric field components $E_x$ and $E_y$ on Earth's surface is that the system is resistive. That is, the approximation $GIC(t)$  $=$ $a_oE_x(t) + b_oE_y(t)$ (Model 1) is used, where $a_o$ and $b_o$ are frequency independent network parameters.  We provide the first test of this assumption using GIC measurements made in a 187 kV transformer connected to a $\sim$100~km power line in Memanbetsu, Japan and geoelectric field measurements made at the Memanbetsu Magnetic Observatory $\sim$9~km away.  We compute frequency dependent network parameters $a(\omega)$ and $b(\omega)$ from data in Model 2, $GIC(\omega)$ $=$ $a(\omega)E_x(\omega) + b(\omega)E_y(\omega)$, and find that they are frequency dependent and that this model provides a significantly better estimates of the measured GIC. A third model, which uses the electric field estimated from a transfer function convolved with the geomagnetic field, provides significantly better estimates to Model~2. It is suggested that the measurement-derived frequency dependence of the system parameters is explained by spatial variations in the spectra of the geoelectric field rather than by an actual frequency dependence of the system parameters.
\end{abstract}

@@

Geomagnetically induced currents (GICs) are electric currents in conducting systems due to electric fields near Earth's surface. Time-varying electric and magnetic fields at Earth's surface are driven by time variations in ionospheric currents on time scales on the order of hours \citep{Ohtani2000} and the movement of Earth's surface relative to slow-varying current systems in Earth's ionosphere that are near stationary relative to the Sun \citep{Stening2013}. Of particular interest are currents induced in electric power systems as they can lead to system degradation, disruption, and failure \citep{Albertson1993,NERC2012}. Accurate estimation of GICs magnitudes is important for power system design, retrospective analysis, and mitigation of the impacts of space weather on power systems \citep{Molinski2002,Thomson2010,NERC2012,Gaunt2014}.

A common assumption made in estimates of GIC using either measured or estimated values of the horizontal geoelectric field at Earth's surface, $\mathbf{E}$, is that the power system is resistive or quasi-DC in the sense that the relationship $GIC(t) = a_oE_x(t) + b_oE_y(t)$ holds \citep{Albertson1981,Lehtinen1985}. In this case, estimates of the coefficients $a_o$ and $b_o$ can be easily made from the data using linear regression, when contemporaneous measurements of GIC and $\mathbf{E}$ are available, or by using information about the connectivity of the power lines and values of the conductor and transformer resistances using DC circuit methods \citep[e.g.][]{Boteler2014a,Boteler2014b}. This quasi-DC assumption has been used or stated in many GIC-related works, for example, \citet{Pulkkinen2007,Wik2008,Pulkkinen2010,Ngwira2011,Horton2012,Viljanen2012,Overbye2012,Marshall2013,Liu2014,Zheng2014,Watari2015,Bonner2017}.

@@

In this work, we use a unique dataset in which measurements $\mathbf{E}$ and the surface horizontal geomagnetic field, $\mathbf{B}$, were made near a site where GICs were being measured in an electric power system.  We first estimate the system parameters $a_o$ and $b_o$ from data using conventional methods, which assume that they are frequency independent. Then, we use a model in which the system parameters are allowed to be frequency dependent, compute the frequency dependence, and compare this model with the traditional frequency independent model.

@@

The 1-second-cadence GIC data used span a total of 35 days across 10 continuous intervals that start and end at midnight Japan Standard Time (JST): 2006-04-03 -- 2006-04-10, 2006-07-25 -- 2006-07-29, 2006-08-05 -- 2006-08-08, 2006-08-19 -- 2006-08-21, 2006-11-07 -- 2006-11-12, 2006-11-28 -- 2006-11-29, 2006-12-01, 2006-12-14 -- 2006-12-15, 2007-11-18 -- 2007-11-20, and 2007-11-22 -- 2007-11-22. These data have been presented previously in \citet{Watari2009} and \cite{Watari2015}. These time intervals are those for which GIC data was made available with the exception of three days for which there were obvious problems with either the electric field or GIC measurements. The GIC dataset includes 1~s raw data and 1~Hz low-pass-filtered raw measurements. The results presented here are for the 1~Hz low-pass-filtered GIC values

One-second-cadence $\mathbf{E}$ and $\mathbf{B}$ measurements from Memambetsu Magnetic Observatory were obtained from the Japan Meteorological Agency data portal.

@@

The only modification made to the magnetic field measurements was the replacement of 36 data points in all components on 2006-08-06 with linearly interpolated values. The 36 values for $B_x$ and $B_y$ were the same and much larger than the surrounding values.

The GIC data contains non-physical spikes that are followed by a relaxation time of 40-80 s. These spikes were removed by identifying times $t_s$ when the change in 1~s $|GIC|$ measurements was greater than $0.05$ Amps and then data in a window $GIC(t_s-2)$ through $GIC(t_s+100)$ were replaced using a linear interpolation of the values at $GIC(t_s-3)$ and $GIC(t_s + 101)$. In the example interval shown in Figure~\ref{sample}, 493 spikes were removed. For all of the time intervals considered, the fraction of interpolated GIC values ranged from 1--4\%.

@@

\noindent
the matrix equation $\underline{\mathbf{GIC}} = \underline{\mathbf{E}}\cdot\mathbf{p}$ is created, where $\underline{\mathbf{GIC}}$ is a zero mean 864000x1 matrix of 1 day observations with rows of $GIC(t)$; $\mathbf{p} = [a_o,b_o]^T$, and $\underline{\mathbf{E}}$ is a 86400x2 matrix with rows of $[E_x(t), E_y(t)]$ observations with each column having zero mean. The values of $\mathbf{p}$ are then obtained using linear regression. The values of $a_o$ and $b_o$ determined in this way correspond to those that minimize the sum-of-squares of $GIC(t)-G_o(t)$ over a 1-day interval. \citep[][provided the mathematically equivalent closed-form equations.]{Pulkkinen2007}. The notation $\overline{a}_o$ and $\overline{b}_o$ is used for the average of the $a_o$ and $b_o$ values computed using the 35 unique 1-day window segments of available data.

@@

\begin{linenomath*}
\begin{equation}
G_E(\omega) = a(\omega)E_x(\omega) + b(\omega)E_y(\omega)
\label{model2}
\end{equation}
\end{linenomath*}
\noindent
and are computed using contemporaneous measurements of $GIC(t)$ and $\mathbf{E}(t)$ using the method described later in this section. Note that fourier transformed variables are indicated by explicitly listing their dependence on $\omega$.

In Model~3, the parameters $a(\omega)$ and $b(\omega)$ from Model~2 are used, but with an electric field $\mathbf{E}'$ computed using the measured magnetic field and a transfer function $\boldsymbol{Z}$: $\mathbf{E}'(\omega) \equiv \boldsymbol{Z}(\omega)\mathbf{B}(\omega)$:

@@

\noindent
where, explicitly, the electric field components are $E'_x$ = $Z_{xx}B_x + Z_{xy}B_y$ and $E'_y = Z_{yx}B_x + Z_{yy}B_y$. 
That is, instead of using the measured electric field directly as in Model 2, a transfer function that relates $\mathbf{E}$ to $\mathbf{B}$ is used to provide an estimate of the electric field based on the magnetic field. In this work, $\boldsymbol{Z}$ was calculated using the method described later in this section.

@@

\noindent
In this model, the transfer function components $z_x$ and $z_y$ are determined directly from measurements of $GIC(t)$ and $\mathbf{B}(t)$, and the model is expected to produce predictions of $GIC(t)$ that are similar to that obtained from Model 3 because they both use the same input variable.

Estimates of $a(\omega)$ and $b(\omega)$ are made at evaluation frequencies, $\omega_e \equiv 2\pi f_e$, by performing a least squares regression using Equation~\ref{model2} and the set $G_E(\omega)$ and $\mathbf{E}(\omega)$ values near $f_e$ from the fourier transform of 24 hours of 1-s cadence data. The values of $f_e$ start at $0.25$ Hz and decrease by a factor of $\sqrt{2}$; if $f_e$ value is not a multiple of $1/86400$, the nearest multiple is chosen. The regression for $f_e$ uses all $f$ values in a window with a width of twice the separation from the next lower value of $f_e$ and the given $f_e$.

To convert the frequency domain transfer functions $a(\omega)$ and $b(\omega)$ estimated at the frequencies $\omega_e$ to impulse responses $h_a(\tau)$ and $h_b(\tau)$, their real and imaginary components are linearly interpolated onto a 1 Hz frequency grid and then inverse fourier transformed.

@@

%It is expected that the time and frequency domain methods should give similar results because the transfer functions in the time- and frequency-domains are mathematically related \citep{Schoukens2004,Ljung2007}, although differences may exist as a result of implementation choices or roundoff \citep{Ljung2004}.  For example, regarding implementation, in the time domain method we choose a finite value of $N_c$ and $N_a$ and in the frequency domain method, we choose the set of evaluation frequencies $\omega_e$, an averaging window function and width, and $\omega$ values (which depend on the length of the time series used to calculate the spectra).

%Because the frequency domain computations are significantly faster, the results presented here are based on the frequency domain results and only spot checks were performed to verify that the time- and frequency-domain methods gave consistent results.

Transfer function parameters were computed for each of the 35 1-day segments of available data. The average transfer function magnitude and phase was obtained by calculating their values each segment and then averaging the segments values.

There are many statistical methods and variations thereof that have been developed for estimating transfer functions in the magnetotelluric (MT) literature that use additional steps and processing \citep{Egbert1986,Chave1987,Chave1989,Jones1989,Larsen1996,Egbert1997,Eisel2001,Chave2004,Chave2012,Chave2017} beyond the ones used in this work that follow the basic method used in \cite{Sims1971} and \cite{Simpson2005}. Several additional processing methods and steps beyond the basic method that have been used in the MT literature were tested individually by modifying the basic least-squares method used to determine if there was a change in the computed transfer functions and, most relevant for this work, the out-of-sample prediction performance of each model. When using these additional steps, we find that the conclusions made regarding the differences each model's prediction ability do not change, the differences in each model's prediction performance is within the calculated uncertainty of the used method, and the transfer function values are generally within the error bars of the basic method used when the signal to noise ratio is above $\sim$3 (corresponding to a period of $\sim 30$~s). The experiments in this regard are discussed in the Appendix.

%The objective of these works is typically the estimation of a robust and low bias transfer function.

%In the results presented, we did not use a robust statistical method or attempt to correct for bias in estimates as our primary interest is comparing the ability of different models to predict the data (in a mean-squared-error sense) rather than estimating the model parameters. As discussed and evaluated in~\cite{Weigel2017}, a model developed using robust and bias correction methods will not necessarily be an optimal prediction model (when the sum-of-squares error is used as a prediction metric).

@@

The out-of-sample prediction metrics for each model on day $D$ were determined by computing the average model parameters using the remaining 34 days of data, using these averaged parameters to generate a prediction time series on day $D$. This process was repeated to create a set of 35 out-of-sample prediction metrics for each model.

@@

The average PE and MSE ratios were computed by averaging the 35 out-of-sample PE and MSE ratios; the average signal-to-noise ratio at a given frequency is the average of the out-of-sample signal-to-noise ratios at that frequency. The 95\% confidence interval, CI, on the PE and MSE ratio was calculated using 1000 bootstrap samples of the out-of-sample prediction metrics. The typical difference between the limits when the confidence intervals were computed assuming the values were Gaussian-distributed was $\sim$2\%.

\begin{table}
\caption{Out-of-sample model performance metrics for models $m=1-4$. MSE$_m$/MSE$_1$ is the ratio of the MSE for model $m$ to that of Model 1. The PE and MSE ratios are the average of 35 values and the confidence interval, CI, was determined from 1000 bootstrap samples of these 35 values.}
\centering
\begin{tabular}{l c c c c}
\hline
$m$\hspace{1em} Model & PE & 95\% CI & MSE$_m$/MSE$_1$ & 95\% CI\\
\hline
1\hspace{1em} $G_o(t) = a_oE_x(t) + b_oE_y(t)$ & 0.35 & [0.25, 0.45] & 1.0 & \\
2\hspace{1em} $G_E(t) = a(\omega)E_x(t) + b(\omega)E_y(t)$ & 0.60 & [0.54, 0.65] & 1.8 & [1.6, 2.0]\\
3\hspace{1em} $G_{E'}(t) = a(\omega)E'_x(t) + b(\omega)E'_y(t)$ & 0.78 & [0.76, 0.80] & 3.1 & [2.8, 3.4]\\
4\hspace{1em} $G_{B}(t) = Z_xB_x(t) + Z_yB_y(t)$ & 0.83 & [0.80, 0.85] & 4.6 & [3.8, 5.3]\\
\hline
\end{tabular}
\end{table}

Figure~\ref{predictions} shows the predictions for each model in a 1-day interval selected because the value of the prediction efficiency for each model was in the same order as that shown in Table~1, i.e., Model 1 has the lowest PE and Model 4 has the lowest.

The primary conclusions of this work follow from the results shown in Table~1: the frequency dependent model, Model~2, is significantly better at predicting GIC than the frequency independent model, Model~1; Model~3, which uses an estimate of the electric field based on magnetic field measurements, provides significantly better estimates of GIC than Model~2; Model~4, in which a transfer function that connects GIC to $\mathbf{B}$ was derived directly from the data, provides significantly better estimates of GIC than Model~3.

In terms of statistical significance, using $10^5$ bootstrap samples, the null hypothesis that the PE of Model~2 is equal to that of Model~1 can be rejected at a significance level of less than $10^{-5}$, with the same result for Model~3 vs. Model~2. The null hypothesis that the PE of Model~4 is equal to that of Model~3 can be rejected at a significance level of $10^{-2}$.

Although inferior, the Model~1 does have a good prediction performance - the PE is 0.35 and the data/model correlation coefficient of 0.64. The top panel of  Figure~\ref{predictions} shows that its predicted GIC generally follows the overall trend of the measured GIC.

%In a similar way that $|\mathbf{B}|(t)$ or $d|\mathbf{B}|(t)/dt$ are often good predictors of $GIC(t)$ \todo{Pierre: Any suggestions?} {\color{red} \textbf{[REFS]}} despite the fact that physically $\mathbf{E}$ drives GIC and $\mathbf{E}$ is related to $\mathbf{B}$ through a frequency dependent transfer function, Model 1 is a fair predictor of GICs, but is a poor approximation relative to Models 2 and 3.

Figure~\ref{SN} shows the average signal-to-noise (i.e., signal to prediction error) ratios as a function of period for the four models. At each evaluation frequency, the SN ratio is the ratio of the average of the SN ratios for each of the 35 segments and the error bars represent 95\% CIs based on 1000 bootstrap samples of the segment SN ratios at each period. The spectra for the signal and prediction error for each of the segments at a given evaluation frequency was obtained by averaging the raw spectra using values between the next higher and lower evaluation frequency.

@@

Below $600$~s (10~min), Model~2 has the largest SN ratio. This result is somewhat visible in the sample interval shown in Figure~\ref{predictions} where the high frequency fluctuations in the prediction error for Model~2 are the smallest among the models. For these periods, the predictive ability of each model is in the order that may be expected based on the data used to derive the model: Model~2, which uses direct measurements of the electric field, has a higher SN ratio than Model~3, which uses an indirect estimate of the electric field. Model~4 has a slightly higher SN than Model~3, and this is expected because they both use the same input of the magnetic field, but Model~3 has an additional source of uncertainty from the parameters in $\mathbf{Z}$. Above $600$~s, the ordering of the SN is not consistent with these expectations and is possibly explained by two factors (1) the measured GIC is based on an integrated electric field along the power line, and the electric field used in Model~2 is a point measurement - the result is that implicit in the model is an assumption that the electric field is uniform over the length of the line on which GIC is measured and (2) the error associated with this assumption has less of an influence on the errors in a model that uses the point measurements of the magnetic field. This issue is addressed in the following section.

@@

The average parameters and 95\% CI for Model 1 are $a_o = $ $73 [62,64]$ A/(V/km) and $b_o =$ $-72.0 \pm [-98,-45]$ A/(V/km) and the distribution of the parameters for the segments is shown in Figure~\ref{histogram}. (Results are similar if 1-minute averages are used; \cite{Watari2015} found $a_o=38.1$ A/(V/km) and $b_o=-7.4$ A/(V/km) for the interval 2006-12-14--2006-12-15 JST time interval).

@@

The average parameters for Models~2 and~4 are shown in Figure~\ref{H}--\ref{Phi}. The error bars at each time step or period is the standard error of the 35 values used to compute the average. The transfer function magnitudes were calculated by linear interpolation of the transfer function magnitudes at the evaluation frequencies (instead of linearly interpolating the real and imaginary components and then computing the magnitude).

Figure~\ref{H} shows that the timescale of variations of an impulse in $\mathbf{E}$ for Model~2 and $\mathbf{B}$ for Model~4 are similar and are different from zero at more than one time value (the impulse response for Model~3 is similar to that of Model~2 and is not shown). By definition, the impulse response for Model~1 is zero except at $t=0$. Note that the sign of the impulse response at $t=0$ is opposite for Model~1 and Model~2 for the $E_y$ coefficient (at $t=0$, $b_o < 0$ while $h_b(t=0)>0$). An explanation for the acausality of the impulse response is given in \cite{Egbert1992,Kelbert2017}.

>>>>

\begin{keypoints}
\item Power system parameters derived from a time series of GIC and geoelectric field measurements are frequency dependent.
\item A GIC prediction model with frequency dependent system parameters significantly outperforms a frequency independent model.
\item A model derived using GIC/geomagnetic field data provides better predictions than a GIC/geoelectric field-based model.
\end{keypoints}

\begin{abstract}
A common assumption used when estimating geomagnetically induced currents (GICs) in a power system given a nearby direct measurement or indirect estimate of the horizontal geoelectric field components $E_x(t)$ and $E_y(t)$ on Earth's surface is that the system is resistive. That is, the approximation $GIC(t)$  $=$ $a_oE_x(t) + b_oE_y(t)$ (Model 1) is used, where $a_o$ and $b_o$ are frequency independent network parameters.  A first test of this assumption is made using GIC measurements made in a 187 kV transformer connected to a $\sim$100~km power line in Memanbetsu, Japan and geoelectric field measurements made at the Memanbetsu Magnetic Observatory $\sim$9~km away.  Frequency-dependent network parameters $a(\omega)$ and $b(\omega)$ computed from data in Model 2, $GIC(\omega)$ $=$ $a(\omega)E_x(\omega) + b(\omega)E_y(\omega)$,  are shown to be frequency dependent and this model provides a significantly better estimates of the measured GIC over Model~1. A third model, which uses the electric field estimated by convolving a magnetotelluric transfer function with the geomagnetic field, provides significantly better estimates over Model~2. It is suggested that the measurement-derived frequency dependence of the system parameters is explained by spatial variations in the spectrum of the geoelectric field along the line in which GIC is measured rather than by an actual frequency dependence of the system parameters.
\end{abstract}

@@

Geomagnetically induced currents (GICs) are electric currents in conducting systems that are due to electric fields near the Earth's surface. Time-varying electric and magnetic fields at Earth's surface are driven by time variations in ionospheric currents on time scales on the order of hours \citep{Ohtani2000} and the movement of Earth's surface relative to slow-varying current systems in Earth's ionosphere that are near stationary relative to the Sun \citep{Stening2013}. Of particular interest are currents induced in electric power systems because they can lead to system degradation, disruption, and failure \citep{Albertson1993,NERC2012,Gaunt2014}. Accurate estimation of GIC magnitudes is important for power system design, retrospective analysis, and mitigation of the impacts of space weather on power systems \citep{Molinski2002,Thomson2010,NERC2012,Gaunt2014}.

A common assumption made in estimates of GIC in an electric power system using either a measured or estimated time series of the horizontal geoelectric field at Earth's surface, $\mathbf{E}(t)$, is that the power system is resistive or quasi-DC in the sense that the relationship $GIC(t) = a_oE_x(t) + b_oE_y(t)$ holds \citep{Albertson1981,Lehtinen1985}. In this case, estimates of the coefficients $a_o$ and $b_o$ can be easily made from the data using linear regression when contemporaneous measurements of $GIC(t)$ and $\mathbf{E}(t)$ are available, or by using information about the connectivity of the power lines and values of the conductor and transformer resistances using DC circuit methods \citep[e.g.,][]{Boteler2014a,Boteler2014b}. This quasi-DC assumption has been used or stated in many GIC-related works, for example, \citet{Pulkkinen2007,Wik2008,Pulkkinen2010,Ngwira2011,Horton2012,Viljanen2012,Overbye2012,Marshall2013,Liu2014,Zheng2014,Watari2015,Bonner2017}.

@@

In this work, we use a unique dataset in which time series measurements of $\mathbf{E}$ and the surface horizontal geomagnetic field, $\mathbf{B}$, were made near a site where GIC was being measured in an electric power system.  We first estimate the system parameters $a_o$ and $b_o$ from data using conventional methods, in which they are modeled as frequency independent. Then, we use a model in which the system parameters are allowed to be frequency dependent, compute the frequency dependence, and compare this model's ability to represent GIC with the traditional frequency-independent model.

@@

The 1-second-cadence GIC data used span a total of 35 days across 10 intervals that start and end at midnight Japan Standard Time (JST): 2006-04-03 -- 2006-04-10, 2006-07-25 -- 2006-07-29, 2006-08-05 -- 2006-08-08, 2006-08-19 -- 2006-08-21, 2006-11-07 -- 2006-11-12, 2006-11-28 -- 2006-11-29, 2006-12-01, 2006-12-14 -- 2006-12-15, 2007-11-18 -- 2007-11-20, and 2007-11-22 -- 2007-11-22. These data have been presented previously in \citet{Watari2009} and \cite{Watari2015}. These time intervals are those for which GIC data was made available with the exception of three days for which there were obvious problems with either the electric field or GIC measurements. The GIC dataset includes 1~s raw data and 1~Hz low-pass-filtered raw measurements. The results presented here are for the 1~Hz low-pass-filtered GIC measurements

One-second-cadence $\mathbf{E}(t)$ and $\mathbf{B}(t)$ measurements from Memambetsu Magnetic Observatory were obtained from the Japan Meteorological Agency data portal.

@@

The only modification made to the magnetic field measurements was the replacement of 36 data points on 2006-08-06 with linearly interpolated values. The 36 values for $B_x$ and $B_y$ were the same and much larger than the surrounding values.

The GIC data contains non-physical spikes that are followed by a relaxation time of 40-80 s. These spikes were removed by identifying times $t_s$ when the change in the 1~s absolute value of $GIC(t)$ was greater than $0.05$~Amps and then values in a window $GIC(t_s-2)$ through $GIC(t_s+100)$ were replaced using a linear interpolation of the values at $GIC(t_s-3)$ and $GIC(t_s + 101)$. In the example interval shown in Figure~\ref{sample}, 493 spikes were removed. For all of the time intervals considered, the fraction of interpolated GIC values ranged from 1--4\%.

@@

\noindent
the matrix equation $\underline{\mathbf{GIC}} = \underline{\mathbf{E}}\cdot\mathbf{p}$ is solved in a least squares sense, where $\underline{\mathbf{GIC}}$ is a zero mean 86400x1 matrix of 1 day observations with rows of $GIC(t)$; $\mathbf{p} = [a_o,b_o]^T$, and $\underline{\mathbf{E}}$ is a 86400x2 matrix with rows of $[E_x(t), E_y(t)]$ with zero mean columns. The values of $a_o$ and $b_o$ determined in this way correspond to those that minimize the sum-of-squares of $GIC(t)-G_o(t)$ over a 1-day interval. \citep[][provided the mathematically equivalent closed-form equations for solving the matrix equation.]{Pulkkinen2007} The notation $\overline{a}_o$ and $\overline{b}_o$ will be used for the average of the $a_o$ and $b_o$ values computed using the 35 1-day window segments of available data.

@@

\begin{linenomath*}
\begin{equation}
G_E(\omega) = a(\omega)E_x(\omega) + b(\omega)E_y(\omega)
\label{model2}
\end{equation}
\end{linenomath*}
\noindent
and are computed using contemporaneous measurements of $GIC(t)$ and $\mathbf{E}(t)$. The method of parameter estimation and generating predictions for Models 2-4 are given at the end of this section and fourier transformed variables are indicated by explicitly listing their dependence on $\omega$.

In Model~3, the parameters $a(\omega)$ and $b(\omega)$ from Model~2 are used, but with an electric field $\mathbf{E}'$ computed using the measured magnetic field, $\mathbf{B}$, and a transfer function $\boldsymbol{Z}$: $\mathbf{E}'(\omega) \equiv \boldsymbol{Z}(\omega)\mathbf{B}(\omega)$ and

@@

\noindent
where, explicitly, the electric field components are $E'_x$ = $Z_{xx}B_x + Z_{xy}B_y$ and $E'_y = Z_{yx}B_x + Z_{yy}B_y$. 
That is, instead of using the measured electric field directly as in Model 2, a magnetotelluric transfer function that relates $\mathbf{E}$ to $\mathbf{B}$ is used to provide an estimate of the electric field based on the magnetic field.

@@

\noindent
In this model, the transfer function components $z_x$ and $z_y$ are determined directly from measurements of $GIC(t)$ and $\mathbf{B}(t)$ and no electric field measurements are used; this model is expected to produce predictions of GIC that are similar to that obtained from Model 3 because they both use the same input variable.

Estimates of $a(\omega)$ and $b(\omega)$ are made at evaluation frequencies, $\omega_e \equiv 2\pi f_e$, by performing a ordinary least squares regression (OLS) using Equation~\ref{model2} and a set $G_E(\omega)$ and $\mathbf{E}(\omega)$ values near $f_e$ from the DFT (discrete fourier transform) of 24 hours of 1-s cadence data. The 28 values of $f_e$ start at $0.25$ Hz and decrease by a factor of $\sqrt{2}$ and end at $1/43200$~Hz; if the $f_e$ value computed in this way is not an integer multiple of $1/86400$, the nearest smaller integer multiple is used. The regression for each $f_e$ uses all DFT frequencies with values in a window with a width of twice the separation from the next lower value of $f_e$ and the given $f_e$ with the exceptions that the lowest evaluation window has a lower bound of $1/86400$~Hz and highest evaluation window has an upper bound of $0.5$~Hz.

@@

Transfer function parameters were computed for each of the segments. The average transfer function magnitude and phase at a given evaluation frequency was obtained by averaging the segment values at that evaluation frequency.

To convert the frequency domain transfer function parameters estimated at the evaluation frequencies to impulse responses, their real and imaginary components are linearly interpolated individually onto the original DFT frequency grid (with spacing of $1/86400$~Hz), and then the inverse DFT is computed. The interpolated transfer function parameters were set to zero at frequencies higher than $0.25$~Hz and the transfer function estimate at $f=0$ was set to zero before interpolation.

Prediction time series were calculated by convolution of the interpolated transfer function parameters with the DFT of the 24-hr long and 1~s cadence model inputs and computing the inverse DFT (e.g, for Model~2, $G_E(f)=a^I(f)E_x(f)+b^I(f)E_y(f)$ was calculated, where the superscript $I$ indicates the interpolated value, and then the inverse DFT of $G_E(f)$ was calculated to give the prediction $G_E(t)$ on a 1~s time grid).

Several modifications to this basic procedure were considered. Robust regression \citep{Egbert1986} with a Huber weighting constant of $1.345$ and an upper limit of 50 iterations and bi-square weighting with a constant of $4.685$ (both constants provide 95\% efficiency for normally distributed errors), both with no hard-cutoff step; time domain pre-whitening and tapering; and weighting the frequencies around each evaluation frequency using Parzen window weights. When using these additional steps individually, we find that the conclusions made regarding the differences each model's prediction ability do not change because the differences in each model's prediction performance are within the calculated uncertainty of the used method. Transfer function parameter values found using these modifications are generally within the error bars of those for the used procedure when the signal to noise ratio is above $\sim$3 (corresponding to a period of $\sim 30$~s; At periods below $\sim$30~s, the electric field measurements are significantly influenced by the instrument response characteristics~\citep{Fujii2015}, aliasing, and the computed values are also expected to be significantly biased due to low signal to noise ratio.)

@@

The out-of-sample prediction metrics for each model on day $D$ were determined by computing the average model parameters using the remaining 34 days of data and then using these averaged parameters to generate a prediction time series and associated metrics on day $D$. This process was repeated to create a set of 35 out-of-sample prediction metrics for each model. In calculating the metrics, the first and last 10 minutes of the day were excluded to reduce the effect of transients that exist in the prediction at the start of the interval for the amount of time that the causal part of the impulse response is non-zero and at the end for the amount of time that the acausal part of the impulse response is non-zero.

@@

The average PE and MSE ratios were computed by averaging the out-of-sample PE and MSE ratios; the average signal-to-noise ratio at a given frequency is the average of the out-of-sample signal-to-noise ratios at that frequency. The 95\% confidence interval, CI, on the PE and MSE ratio was calculated using 1000 bootstrap samples of the out-of-sample prediction metrics. (The typical difference between the limits when the confidence intervals were computed assuming the values were Gaussian-distributed was $\sim$2\%.) The spectra for the signal and prediction error for each of the segments at a given evaluation frequency was obtained by averaging the raw spectra using the same frequency windows used for the regression.

\begin{table}
\caption{Out-of-sample model performance metrics for models $m=1-4$. MSE$_m$/MSE$_1$ is the ratio of the MSE for model $m$ to that of Model 1. The PE and MSE ratios are the average of 35 values and the confidence interval, CI, was determined from 1000 bootstrap samples.}
\centering
\begin{tabular}{l c c c c}
\hline
$m$\hspace{1em} Model & PE & 95\% CI & MSE$_m$/MSE$_1$ & 95\% CI\\
\hline
1\hspace{1em} $G_o(t) = a_oE_x(t) + b_oE_y(t)$ & 0.35 & [0.25, 0.45] & 1.0 & \\
2\hspace{1em} $G_E(t) = a(\omega)E_x(t) + b(\omega)E_y(t)$ & 0.60 & [0.54, 0.65] & 1.8 & [1.6, 2.0]\\
3\hspace{1em} $G_{E'}(t) = a(\omega)E'_x(t) + b(\omega)E'_y(t)$ & 0.78 & [0.76, 0.80] & 3.1 & [2.8, 3.4]\\
4\hspace{1em} $G_{B}(t) = Z_xB_x(t) + Z_yB_y(t)$ & 0.83 & [0.80, 0.85] & 4.6 & [3.8, 5.3]\\
\hline
\end{tabular}
\end{table}

Figure~\ref{predictions} shows the predictions for each model in a 1-day interval selected because the value of the prediction efficiency for each model was in the same order as that shown in Table~1, i.e., Model 1 has the lowest PE and Model 4 has the highest.

The primary conclusions of this work follow from the results shown in Table~1: the frequency dependent model, Model~2, is significantly better at predicting GIC than the frequency independent model, Model~1; Model~3, which uses an estimate of the electric field based on magnetic field measurements, provides significantly better estimates of GIC than Model~2; Model~4, in which a transfer function that connects GIC to $\mathbf{B}$ was derived directly from the data, provides better estimates of GIC than Model~3.

To determine the statistical significance, $10^5$ bootstrap averages of the PEs of each model were created and their differences were computed. From this, the null hypothesis that the PE of Model~2 is equal to that of Model~1 can be rejected at a significance level of less than $10^{-5}$ because the bootstrap average of Model~2 was not found to be lower than that for Model~1. The null hypothesis that the PE of Model~3 is equal to that of Model~2 can also be rejected at the same significance level. The null hypothesis that the PE of Model~4 is equal to that of Model~3 can be rejected at a significance level of $10^{-2}$.

%Although inferior, the Model~1 does have a good prediction performance - the PE is 0.35 and the data/model correlation coefficient of 0.64. The top panel of  Figure~\ref{predictions} shows that its predicted GIC generally follows the overall trend of the measured GIC.

Figure~\ref{SN} shows the average signal-to-noise (i.e., signal to prediction error) ratios as a function of period for the four models. At each evaluation frequency, the SN ratio is the ratio of the average of the out-of-sample SN ratios of the segments and the error bars represent a 95\% CIs based on 1000 bootstrap samples of the segment SN ratios.

@@

Below $600$~s (10~min), Model~2 has the largest SN ratio. This result is somewhat visible in the sample interval shown in Figure~\ref{predictions} where the high-frequency fluctuations in the prediction error for Model~2 are the smallest among the models. For these periods, the predictive ability of each model is in the order that may be expected based on the data used to derive the model: Model~2, which uses direct measurements of the electric field, has a higher SN ratio than Model~3, which uses an indirect estimate of the electric field. Model~4 has a slightly higher SN than Model~3, and this is expected because they both use the same input of the magnetic field, but Model~3 has an additional source of uncertainty from the parameters in $\mathbf{Z}$. Above $600$~s, the ordering of the SN is not consistent with these expectations and is possibly explained by two factors (1) the measured GIC is based on an integrated electric field along the power line, and the electric field used in Model~2 is a point measurement - the result is that implicit in the model is an assumption that the electric field is uniform over the length of the line on which GIC is measured and (2) the error associated with this assumption has less of an influence on the errors in a model that uses the point measurements of the magnetic field. Issues related to this are discussed in the following section.

@@

The average parameters and 95\% CI for Model 1 are $a_o = 73$ $[62,83]$ A/(V/km) and $b_o = -72.0$ $[-98,-45]$ A/(V/km) and the distribution of the parameters for the segments is shown in Figure~\ref{histogram}. (Results are similar if 1-minute averages are used; \cite{Watari2015} found $a_o=38.1$ A/(V/km) and $b_o=-7.4$ A/(V/km) for the interval 2006-12-14--2006-12-15 JST time interval).

@@

The average parameters for Models~2 and~4 are shown in Figure~\ref{H}--\ref{Phi}. The error bars at each time step or period is the standard error of the 35 values used to compute the average.

%Figure~\ref{H} shows that the timescale of variations of an impulse in $\mathbf{E}$ for Model~2 and $\mathbf{B}$ for Model~4 are similar and are different from zero at more than one time value (the impulse response for Model~3 is similar to that of Model~2 and is not shown). By definition, the impulse response for Model~1 is zero except at $t=0$. Note that the sign of the $E_y$ impulse response at $t=0$ for Model~1 is opposite that for Model~2 (at $t=0$, $b_o < 0$ while $h_b(t=0)>0$). (An explanation for the acausality of the impulse response is given in \cite{Egbert1992,Kelbert2017}).

Figure~\ref{Z} shows that the frequency domain transfer functions for Model~2 vary by an order of magnitude on timescales of 60~s to 12~h (that for Model~1 is constant by definition). For this system, GIC more sensitive to $E_y$ than $E_x$ due to $b(\omega) > a(\omega)$, but note that $E_y(t)$ has less variability than $E_x(t)$ - the variance in $E_x(t)$ is $\sim$5x larger than $E_y(t)$ over the entire dataset (this is also visible in Figure~\ref{sample}). The ratio of $|\overline{b}_o|/|\overline{a}_o|$ is approximately 1 for Model~1 and this can be compared with $|b(\omega)|/|a(\omega)|$ calculated using the lines for $|a(\omega)|$ and $|b(\omega)|$ shown in Figure~\ref{Z}. The ratio varies from 2.7 at $T=30$~s to 1.6 at $T=12$~h. At $T=12$~h, the highest period considered, the phases of $a(\omega)$ and $b(\omega)$ shown in Figure~\ref{Phi} differ by 180 degrees and the signs are consistent with $\overline{a}_o$ and $\overline{b}_o$.
